---
title: "Platte_Trash"
author: "Jack Rodenburg"
date: "2023-11-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#```{r}
# Assuming you have a 'Year' or 'Date' column in shannon_data_extend that corresponds to each row in diversity_data
diversity_data#$Time <- shannon_data_extend$Year  # Replace 'Year' with 'Date' or the appropriate time column

# Reshape the data to long format
long_diversity_data <- gather(diversity_data, "Index", "Value", -Time)

# Plotting each diversity index as a separate line
ggplot(long_diversity_data, aes(x = Time, y = Value, color = Index)) +
  geom_line() +
  labs(title = "Diversity Indices Over Time",
       x = "Time",
       y = "Index Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
#```


#```{r}
#pairs shannon diversity index with sample id
shannon_data <- data.frame(sample_id = fish_vegan$sample_id, Shannon_Diversity = shannon_div)

#creates separate columns with segment and date as separate columns
shannon_data_extend <- shannon_data %>%
  separate(sample_id, into = c("Segment", "Year", "Month", "Day"), sep = "-") %>%
  filter(Segment != "NA") %>% 
  mutate(Segment = factor(Segment, levels = c("CP", "ULP", "ELK", "LLP", "PMC"))) %>% 
  mutate(Date = as.Date(paste(Year, Month, Day, sep = "-"), format = "%Y-%m-%d"))

#Calculates each unique shannon diversity for each time point
shannon_avg_time <- shannon_data_extend %>%
  group_by(Date) %>%
  summarize(Avg_Shannon_Diversity = mean(Shannon_Diversity))

#shannon_avg_time

ggplot(shannon_avg_time, aes(x = Date, y = Avg_Shannon_Diversity)) +
  geom_line(color = "black") +
  #geom_point(color = "blue") +
  labs(title = "Shannon Diversity Over Time",
       x = "Date",
       y = "Average Shannon Diversity") +
  theme_minimal()

diversity_data_extend
#```

This chunk smooths out ths above data, average shannon diveristy across months instead of sample periods



```{r}
# Calculate the average Shannon diversity for each month
shannon_avg_monthly <- shannon_data_extend %>%
  group_by(Year, Month) %>%
  summarize(Avg_Shannon_Diversity = mean(Shannon_Diversity)) %>%
  ungroup() %>%
  mutate(Month = as.numeric(Month),
         YearMonth = as.Date(paste(Year, Month, "01", sep = "-"), format="%Y-%m-%d")) %>%
  arrange(Year, Month)

# Plot the average Shannon diversity for each month, with only years on the x-axis
ggplot(shannon_avg_monthly, aes(x = YearMonth, y = Avg_Shannon_Diversity)) +
  geom_line() +  # Add lines connecting the points
  geom_point(aes(color = Year)) +  # Points colored by year
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +  # Format x-axis labels to show only years
  labs(title = "Average Monthly Shannon Diversity Over Years",
       x = "Year",
       y = "Average Shannon Diversity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))  # Rotate x-axis labels for readability
```
This chunk calculates Shannon diverisity for each half year

```{r}
# Calculate the average Shannon diversity for each half-year
shannon_avg_halfyearly <- shannon_data_extend %>%
  mutate(Half = ifelse(Month %in% 1:6, "H1", "H2"),
         YearHalf = paste(Year, Half, sep = "-")) %>%
  group_by(Year, Half, YearHalf) %>%
  summarize(Avg_Shannon_Diversity = mean(Shannon_Diversity), .groups = 'drop') %>%
  mutate(YearHalfDate = as.Date(paste(Year, ifelse(Half == "H1", "06", "12"), "30", sep = "-"))) %>%
  arrange(YearHalfDate)  # Ensure correct ordering for plotting lines

# Plot the average Shannon diversity for each half-year, connecting all points
ggplot(shannon_avg_halfyearly, aes(x = YearHalfDate, y = Avg_Shannon_Diversity)) +
  geom_line() +  # Connect all points with lines
  geom_point(aes(color = as.factor(Year))) +  # Points colored by year
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +  # Format x-axis labels to show only years
  labs(title = "Average Half-Yearly Shannon Diversity Over Years",
       x = "Year",
       y = "Average Shannon Diversity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for readability

# Execute the plot
print(shannon_avg_halfyearly)
```

This chunk calculates average shannon diversity for each year
```{r}
# Ensure Year is treated as numeric for the plotting
shannon_avg_yearly <- shannon_data_extend %>%
  mutate(Year = as.numeric(as.character(Year))) %>%
  group_by(Year) %>%
  summarize(Avg_Shannon_Diversity = mean(Shannon_Diversity), .groups = 'drop')

# Define the breaks for the x-axis
year_breaks <- shannon_avg_yearly$Year

# Plot the average Shannon diversity for each year with connected lines
ggplot(shannon_avg_yearly, aes(x = Year, y = Avg_Shannon_Diversity)) +
  geom_line() +  # Connect points with lines
  geom_point() +
  #geom_point(aes(color = as.factor(Year))) +  # Points colored by year
  scale_x_continuous(breaks = year_breaks) +  # Set x-axis breaks to show each year
  labs(title = "Average Yearly Shannon Diversity Over Years",
       x = "Year",
       y = "Average Shannon Diversity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for readability
```



This chunk calculates the average shannon diversity for each location across years 
```{r}
#calculates each shannon diversity for each unique location
shannon_avg_location <- shannon_data_extend %>%
  group_by(Segment) %>%
  summarize(Avg_Shannon_Diversity = mean(Shannon_Diversity))

#plots Shannon diversity for each location
ggplot(shannon_avg_location, aes(x = Segment, y = Avg_Shannon_Diversity)) +
  geom_line(color = "black") +
  geom_point(color = "blue") +
  labs(title = "Shannon Diversity Across Locations",
       x = "Segment",
       y = "Average Shannon Diversity") +
  theme_minimal()
```

This chunk finds the maximum Shannon diversity experienced by each place
#i want to place a label on each data point indicating the month,year max diversity was acheived#

```{r}
# Prepare the data with Year and Shannon Diversity
shannon_yearly_max <- shannon_data_extend %>%
  mutate(Year = as.numeric(as.character(Year))) %>%
  group_by(Year) %>%
  summarize(Max_Shannon_Diversity = max(Shannon_Diversity)) %>%
  ungroup()  #to remove the grouping

shannon_yearly_max
year_breaks <- shannon_avg_yearly$Year


# Plot the data
ggplot(shannon_yearly_max, aes(x = as.numeric(Year), y = Max_Shannon_Diversity)) +
  geom_line(group = 1, color = "blue") +  # Line plot
  geom_point(color = "red") +  # Add points
  scale_x_continuous(breaks = year_breaks) +
  #scale_x_continuous(breaks = scales::pretty_breaks(n = 10), minor_breaks = NULL) +  # Pretty breaks for years
  labs(title = "Maximum Shannon Diversity Per Year",
       x = "Year",
       y = "Maximum Shannon Diversity") +
  theme_minimal()
```


```{r}
# Prepare the data with Year and Shannon Diversity
shannon_yearly_med <- shannon_data_extend %>%
  group_by(Year) %>%
  summarize(Median_Shannon_Diversity = median(Shannon_Diversity)) %>%
  ungroup()  # to remove the grouping

shannon_yearly_med
# Define the breaks for the x-axis
year_breaks <- shannon_avg_yearly$Year

# Plot the data
ggplot(shannon_yearly_med, aes(x = as.numeric(Year), y = Median_Shannon_Diversity)) +
  geom_line(group = 1, color = "blue") +  # Line plot
  geom_point(color = "red") +  # Add points
  scale_x_continuous(breaks = year_breaks) +
  #scale_x_continuous(breaks = scales::pretty_breaks(n = 10), minor_breaks = NULL) +  # Pretty breaks for years
  labs(title = "Median Shannon Diversity Per Year",
       x = "Year",
       y = "Median Shannon Diversity") +
  theme_minimal()
```
This calculates average shannon diversity per segment per year. 
```{r}
#Prepare the data by grouping by Segment and Year
shannon_location_year <- shannon_data_extend %>%
  group_by(Segment, Year) %>%
  summarize(Avg_Shannon_Diversity = mean(Shannon_Diversity)) %>%
  ungroup()  # Ungroup to avoid issues in further manipulation

#Plot
ggplot(shannon_location_year, aes(x = Segment, y = Avg_Shannon_Diversity, group = Year, color = as.factor(Year))) +
  geom_line() +  # Add lines
  geom_point(aes(shape = as.factor(Year))) +  # Add points and use different shapes for each year
  scale_color_manual(values = rainbow(length(unique(shannon_location_year$Year)))) +  # Set manual colors if desired
  labs(title = "Average Shannon Diversity Across Locations by Year",
       x = "Location (Segment)",
       y = "Average Shannon Diversity") +
  theme_minimal() +
  theme(legend.title = element_blank())  # Hide the legend title if you wish
```
This chunk will calculate the species richness for each data point collected

```{r}
# Add a Date column to diversity_data based on the matching row number in shannon_data_extend
diversity_data$Date <- shannon_data_extend$Date

# Now plot the Richness for each sample date
ggplot(diversity_data, aes(x = Date, y = Richness)) +
  geom_line() +  # Add lines connecting the points
  geom_point(aes(color = as.factor(format(Date, "%Y")))) +  # Points colored by year
  labs(title = "Species Richness Over Time",
       x = "Date",
       y = "Species Richness") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```
This chunk will calculate the average species richness for each month point collected

```{r}
# Add a year-month column for grouping
diversity_data$YearMonth <- format(diversity_data$Date, "%Y-%m")

# Calculate the average species richness for each month
average_monthly_richness <- diversity_data %>%
  group_by(YearMonth) %>%
  summarize(Average_Richness = mean(Richness, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(YearMonth = as.Date(paste(YearMonth, "01", sep = "-")))  # Convert YearMonth back to a Date object for plotting

# Plot the average species richness per month
ggplot(average_monthly_richness, aes(x = YearMonth, y = Average_Richness)) +
  geom_line() +  # Add lines connecting the points
  geom_point(aes(color = as.factor(year(YearMonth)))) +  # Points colored by year
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +  # Set x-axis breaks to show only years
  labs(title = "Average Monthly Species Richness Over Time",
       x = "Year",
       y = "Average Species Richness") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels for readability

```
```{r}
# Ensure that diversity_data has the Segment information
diversity_data$Segment <- shannon_data_extend$Segment

# Calculate the average species richness for each location (Segment)
richness_avg_location <- diversity_data %>%
  group_by(Segment) %>%
  summarize(Avg_Richness = mean(Richness, na.rm = TRUE)) %>%
  ungroup()

# Plot average species richness for each location
ggplot(richness_avg_location, aes(x = Segment, y = Avg_Richness)) +
  geom_line(color = "black") +
  geom_point(color = "blue") +
  labs(title = "Species Richness Across Locations",
       x = "Segment",
       y = "Average Species Richness") +
  theme_minimal()
```


```{r}
# Ensure that diversity_data has Segment and Year information
diversity_data$Segment <- shannon_data_extend$Segment
diversity_data$Year <- shannon_data_extend$Year

# Calculate the average species richness for each segment per year
richness_location_year <- diversity_data %>%
  group_by(Segment, Year) %>%
  summarize(Avg_Richness = mean(Richness, na.rm = TRUE)) %>%
  ungroup()

# Plot the average species richness across locations by year
ggplot(richness_location_year, aes(x = Segment, y = Avg_Richness, group = Year, color = as.factor(Year))) +
  geom_line() +  # Add lines
  geom_point(aes(shape = as.factor(Year))) +  # Add points and use different shapes for each year
  scale_color_manual(values = rainbow(length(unique(richness_location_year$Year)))) +  # Set manual colors if desired
  labs(title = "Average Species Richness Across Locations by Year",
       x = "Location (Segment)",
       y = "Average Species Richness") +
  theme_minimal() +
  theme(legend.title = element_blank())  # Hide the legend title if you wish
```


```{r}
# Filter out the "ELK" and "PMC" segments
richness_location_year_filtered <- richness_location_year %>%
  filter(Segment != "ELK", Segment != "PMC")

# Plot the average species richness across locations by year, excluding "ELK" and "PMC"
ggplot(richness_location_year_filtered, aes(x = Segment, y = Avg_Richness, group = Year, color = as.factor(Year))) +
  geom_line() +  # Add lines
  geom_point(aes(shape = as.factor(Year))) +  # Add points and use different shapes for each year
  scale_color_manual(values = rainbow(length(unique(richness_location_year_filtered$Year)))) +  # Set manual colors if desired
  labs(title = "Average Species Richness Across Locations by Year (Excluding ELK and PMC)",
       x = "Location (Segment)",
       y = "Average Species Richness") +
  theme_minimal() +
  theme(legend.title = element_blank())  # Hide the legend title if you wish
```

```{r}
# Ensure that diversity_data has Segment and Year information
diversity_data$Segment <- shannon_data_extend$Segment
diversity_data$Year <- shannon_data_extend$Year

# Filter out the "ELK" and "PMC" segments
richness_location_year_filtered <- diversity_data %>%
  filter(Segment != "ELK", Segment != "PMC") %>%
  group_by(Segment, Year) %>%
  summarize(Avg_Richness = mean(Richness, na.rm = TRUE)) %>%
  ungroup()

# Plot the average species richness across years by Segment
ggplot(richness_location_year_filtered, aes(x = Year, y = Avg_Richness, group = Segment, color = Segment)) +
  geom_line() +  # Add lines
  geom_point() +  # Add points
  scale_color_manual(values = rainbow(length(unique(richness_location_year_filtered$Segment)))) +  # Set manual colors for each Segment
  labs(title = "Average Species Richness Across Years by Segment",
       x = "Year",
       y = "Average Species Richness") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```

```{r}
CLUSTER ANALYSIS
```{r}
numeric_data_scaled <- scale(numeric_data)
distance_matrix <- dist(numeric_data_scaled, method = "euclidean")
hc <- hclust(distance_matrix)
clusters <- cutree(hc, k = 1) # for example, creating 5 clusters
plot(hc)

numeric_data_scaled$Cluster <- as.factor(clusters)
long_data <- melt(numeric_data_scaled, id.vars = "Cluster")
#ggplot(long_data, aes(x = Species, y = reorder(sample_id, Cluster), fill = Abundance)) +
  #geom_tile() +
  #scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       #midpoint = 0, limit = c(-3, 3), space = "Lab",
                       #name="Scaled\nAbundance") +
  #theme_minimal() +
  #theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  #labs(x = "Species", y = "Sample", title = "Heatmap of Fish Species Abundance by Cluster")

```

```{r}
platte_fish_cleaned <- platte_fish_cleaned %>%
  mutate(Count = as.numeric(as.character(Count)))

#isolates just the carp
bass_data_avg <- platte_fish_cleaned %>%
  filter(Species %in% c(620, 645, 714, 750, 770))

bass_data_avg <- bass_data_avg %>%
  filter(!is.na(Count))

# Summarize the data to get the total count per year (or per month if needed)
bass_count_yearly <- bass_data_avg %>%
  group_by(Year) %>%
  summarize(Total_Count = sum(Count, na.rm = TRUE)) %>%
  ungroup()

# Plot the catfish count over time
ggplot(bass_count_yearly, aes(x = Year, y = Total_Count)) +
  geom_smooth() +  # Add lines connecting the points
  geom_point(color = "blue") +  # Add points
  scale_x_continuous(breaks = unique(bass_count_yearly$Year)) +  # Specify x-axis breaks to be each year
  labs(title = "Bass Population Over Time",
       x = "Year",
       y = "Total Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```




```{r}
# Summarize the data to get the total count per year for each species
carp_count_yearly_species <- carp_data %>%
  group_by(Year, Species) %>%
  summarize(Total_Count = sum(Count, na.rm = TRUE)) %>%
  ungroup()

# Plot the counts for each species over time
ggplot(carp_count_yearly_species, aes(x = Year, y = Total_Count, color = as.factor(Species))) +
  geom_smooth() +  # Add smooth lines without standard error
  geom_point() +  # Add points
  scale_x_continuous(breaks = unique(carp_count_yearly_species$Year)) +  # Specify x-axis breaks to be each year
  labs(title = "Carp Population by Species Over Time",
       x = "Year",
       y = "Total Count",
       color = "Species") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```{r}
platte_fish_cleaned <- platte_fish_cleaned %>%
  mutate(Count = as.numeric(as.character(Count)))

#isolates just the carp
minnow_data <- platte_fish_cleaned %>%
  filter(Species %in% c(150,152,154,156,158,160,162))




minnow_data <- minnow_data %>%
  filter(!is.na(Count))

# Summarize the data to get the total count per year (or per month if needed)
minnow_count_yearly <- minnow_data %>%
  group_by(Year) %>%
  summarize(Total_Count = sum(Count, na.rm = TRUE)) %>%
  ungroup()

# Plot the catfish count over time
ggplot(minnow_count_yearly, aes(x = Year, y = Total_Count)) +
  geom_line() +  # Add lines connecting the points
  geom_point(color = "blue") +  # Add points
  scale_x_continuous(breaks = unique(minnow_count_yearly$Year)) +  # Specify x-axis breaks to be each year
  labs(title = "Minnow Population Over Time",
       x = "Year",
       y = "Total Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```
```

